---
title: "Data notes -- Diagnoses for Hospital Stays, 2014 to 2016"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Source
The Comprehensive Hospital Abstract Reporting System (CHARS) is a Department of Health (DOH) system which collects record level information on inpatient and observation patient community hospital stays. For more details on CHARS see the DOH CHARS information page [[link](https://www.doh.wa.gov/DataandStatisticalReports/HealthcareinWashington/HospitalandPatientData/HospitalDischargeDataCHARS)].

## Ranking of Diagnoses of Hospital Stays
Two types of ranking were provided: one is based on principal diagnosis that is the condition established after study to be chiefly responsible for occasioning the admission of the patient for care; another is based on  principal diagnosis as well as other diagnosis conditions that coexist at the time of admission, that develop subsequently, or that affect the treatment received and/or the length of stay (up to 24 diagnoses)

## Hospital Stays
The unit of analysis is the hospital discharge (i.e., the inpatient stay), not a person or patient. This means that a person who is admitted to the hospital multiple times in one year will be counted each time as a separate "discharge" from the hospital. Counts are summarized by discharge year. 

## Diagnosis/Condition Categorization
Clinical Classification Software (CCS) and Medical Expenditure Panel Survey (MEPS) Condition Categories are used to categorize diagnoses in CHARS (Appendix).

####  Clinical Classifications Software (CCS) Categories
The Clinical Classifications Software (CCS) for ICD-9-CM [[link](https://www.hcup-us.ahrq.gov/toolssoftware/ccs/ccs.jsp)] and for ICD-10-CM [[link](https://www.hcup-us.ahrq.gov/toolssoftware/ccs10/ccs10.jsp)] categorizes ICD-9-CM and ICD-10-CM diagnosis codes into a manageable number of clinically meaningful categories to facilitate statistical analysis and reporting. The CCS for ICD-10-CM undergoes periodic updates and corrections as data using ICD-10 codes come into greater use. The Version 2018.1 of CCS ICD-10-CM is applied in the report. 

CCS consists of two related classification systems, single-level and multi-level, which are designed to meet different needs. Single-level CCS that aggregates illnesses and conditions into 285 mutually exclusive categories is used in the report. We end up with 237 categories after performing the following steps:

* Maternal and neonatal stays are grouped by Multiple-level CCS  (Multi-CCS-level-1 11: 'Complications of pregnancy; childbirth; and the puerperium' and Multi-CCS-level-1 15: 'Certain conditions originating in the perinatal period');
* The category 259 'Residual codes; unclassified' was excluded;
* The external causes of injury and poisoning (E-codes: E800-E999) for ICD-9-CM and the external causes of morbidity (V, W, X, Y codes: V01-Y99) are not listed in diagnosis conditions in CHARS and the single-level CCS categories 260 and 2601-2621 for external causes codes are excluded.

####  Medical Expenditure Panel Survey (MEPS) Condition Categories
MEPS condition categories group single-level CCS categories into 61 broader categories. This table [[link](http://www.meps.ahrq.gov/mepsweb/data_stats/conditions.shtml)] shows the CCS codes that make up each category. Decisions about collapsing CCS categories were based on maintaining the clinical significance of categories, the ability of MEPS respondents to report the condition accurately, and the frequency of reporting the condition. The category 'Residual codes' (the corresponding CCS category 259 'Residual codes; unclassified') was excluded in the report.

## Caution: Transition from ICD-9-CM to ICD-10-CM/PCS Coding in 2015
On October 1, 2015, the United States transitioned from ICD-9-CM to ICD-10-CM, resulting in the 2014 CHARS data with ICD-9-CM codes, the 2015 CHARS data with both ICD-9-CM (the first three quarters) and ICD-10-CM (the last quarter) codes, and the 2016 CHARS data with ICD-10-CM codes. Every effort was made by AHRQ-HCUP to translate the CCS system to ICD-10-CM without making changes to the CCS assignments for diseases and conditions. However, because of the structure and greater specificity of ICD-10-CM there are some ICD-9-CM conditions that do not map to the same CCS in ICD-10-CM. Thus, it should be cautious when you compare the annual data in the report (use the CCS to ICD-9-CM [[link](https://www.hcup-us.ahrq.gov/toolssoftware/ccs/ccs.jsp)] /ICD-10-CM [[link](https://www.hcup-us.ahrq.gov/toolssoftware/ccs10/ccs10.jsp)] crosswalk as reference). 

## Table of The Files Received from DOH

| File Name (xlsx file) | Description|
| --------------------- | ------------------------------------------------ |
|hospitalization_<ACH name>_Primary_diagnosis_ccs_comb | CCS-based ranking of primary diagnosis of Hospital Stays (ACH and county level) |
|hospitalization_<ACH name>_Primary_diagnosis_MEPS | MEPS-based ranking of primary diagnosis of Hospital Stays (ACH and county level) |
|hospitalization_<ACH name>_All_25_diagnoses_ccs_comb | CCS-based ranking of all 25 diagnoses of Hospital Stays (ACH and county level) |
|hospitalization_<ACH name>_All_25_diagnoses_MEPS | MEPS-based ranking of all 25 diagnoses of Hospital Stays (ACH and county level) |


## Appendix

**Table: Category Labels for CCS and MEPS**
```{r, echo=FALSE}
lb <- read.csv('V:/Staff/JXG4303/CHARS/CCS_HCUP/CCS_ICD9/Single_Level_CCS_2015/dxlabel 2015.csv', skip=4, header=F, stringsAsFactors = F)
names(lb) <- c('CCS_category', 'CCS_label')
lb_meps <- xlsx::read.xlsx('V:/Staff/JXG4303/CHARS/CCS_HCUP/CCS_grouping_MEPS_AHRQ_2007.xlsx', sheetIndex = 1, stringsAsFactors=FALSE)
lb_meps$v1 <- sub(' - ', ':', lb_meps$CCS.Codes)
lb$MEPS_label <- NA
for (i in 1:dim(lb_meps)[1]) {
  lb$MEPS_label[lb$CCS_category %in% eval(parse(text=paste0('c(', noquote(lb_meps$v1[i]),')')))] <- lb_meps$Condition.Category[i]
}
lb$MEPS_label[is.na(lb$MEPS_label)==1] <- 'External causes of injury and poisoning'
knitr::kable(lb, align='l')
```


