---
title: "Data notes for diagnoses for hospital stays, 2014 to 2016"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Source
The Comprehensive Hospital Abstract Reporting System (CHARS) is a Department of Health (DOH) system which collects record level information on inpatient and observation patient community hospital stays. For more details on CHARS see the DOH CHARS information page [[link](https://www.doh.wa.gov/DataandStatisticalReports/HealthcareinWashington/HospitalandPatientData/HospitalDischargeDataCHARS)].

## Ranking of Diagnoses of Hospital Stays
Two types of ranking were provided: one is based on principal diagnosis that is the condition established after study to be chiefly responsible for occasioning the admission of the patient for care; another is based on  principal diagnosis as well as other diagnosis conditions that coexist at the time of admission, that develop subsequently, or that affect the treatment received and/or the length of stay (up to 24 diagnoses)


Inpatient Stays
The unit of analysis in the NIS is the hospital discharge (i.e., the inpatient stay), not a person or patient. This means that a person who is admitted to the hospital multiple times in one year will be counted each time as a separate "discharge" from the hospital. Counts are summarized by discharge year. There were no exclusions applied to the data




When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
