---
title: "Data notes for diagnoses for hospital stays, 2014 to 2016"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Source
The Comprehensive Hospital Abstract Reporting System (CHARS) is a Department of Health (DOH) system which collects record level information on inpatient and observation patient community hospital stays. For more details on CHARS see the DOH CHARS information page [[link](https://www.doh.wa.gov/DataandStatisticalReports/HealthcareinWashington/HospitalandPatientData/HospitalDischargeDataCHARS)].

## Ranking of Diagnoses of Hospital Stays
Two types of ranking were provided: one is based on principal diagnosis that is the condition established after study to be chiefly responsible for occasioning the admission of the patient for care; another is based on  principal diagnosis as well as other diagnosis conditions that coexist at the time of admission, that develop subsequently, or that affect the treatment received and/or the length of stay (up to 24 diagnoses)

## Hospital Stays
The unit of analysis is the hospital discharge (i.e., the inpatient stay), not a person or patient. This means that a person who is admitted to the hospital multiple times in one year will be counted each time as a separate "discharge" from the hospital. Counts are summarized by discharge year. 

## Clinical Classifications Software (CCS) Categories
The Clinical Classifications Software (CCS) for ICD-9-CM [[link](https://www.hcup-us.ahrq.gov/toolssoftware/ccs/ccs.jsp)] and for ICD-10-CM [[link](https://www.hcup-us.ahrq.gov/toolssoftware/ccs10/ccs10.jsp)] categorizes ICD-9-CM and ICD-10-CM diagnosis codes into a manageable number of clinically meaningful categories to facilitate statistical analysis and reporting. CCS consists of two related classification systems, single-level and multi-level, which
are designed to meet different needs. Single-level CCS that aggregates illnesses and conditions into 285 mutually exclusive categories is used in the report.  

Maternal and neonatal stays are grouped by Multiple-level CCS  (Multi-CCS-level-1 11: 'Complications of pregnancy; childbirth; and the puerperium' and Multi-CCS-level-1 15: 'Certain conditions originating in the perinatal period').






## Medical Expenditure Panel Survey (MEPS) Condition Categories

##comparability across years
Every effort was made to translate the CCS system to ICD-10-CM/PCS without making changes to the CCS assignments for diseases and conditions. However, because of the structure and greater specificity of ICD-10-CM there are some ICD-9-CM conditions that do not map to the same CCS in ICD-10-CM. For example; ICD-9-CM code 721.8, Other allied disorders of spine, is assigned to CCS 205, Spondylosis; intervertebral disc disorders; other back problems. The GEMs map this ICD-9-CM code to two ICD-10-CM codes. 

M25.78, Osteophyte, vertebrae, assigned to CCS 212, Other bone disease and musculoskeletal deformities
M48.9, Spondylopathy, unspecified, assigned to CCS 205, Spondylosis; intervertebral disc disorders; other back problems

Even though the code description for M48.9 is more specific than the ICD-9-CM code 721.8, the CCS assignment to CCS 205 was preserved. However, code M25.78 is much more specific than 721.8 and resulted in a different CCS assignment, CCS 212, because osteophyte did not fit into CCS 205. Therefore, when users of the system are comparing ICD-9-CM to ICD-10-CM, it is important to remember that one ICD-9-CM code may map to multiple ICD-10-CM codes, of which some may be closer matches to the ICD-9-CM code/description than others. 





## E code
In ICD-10-CM, the E codes have moved to Chapter 20: External Causes of Morbidity (V01-Y99). The external cause codes are used essentially the same in ICD-10-CM as they are in ICD-9-CM, with a few modifications
No external cause code from ICD-10-CM Chapter 20 is needed if the external cause and intent are included in a code from another chapter, such as T42.3X1A, Poisoning by barbituates, accidental, intentional, initial encounter. This is a new guideline for ICD-10-CM.

ICD9: Supplemental Classification of External Causes of Injury and Poisoning (E-codes, E800-E999),     E code can never be a principal diagnosis
ICD10: Chapter 20: External Causes of Morbidity (V00-Y99) The external causes of morbidity, codes should never be sequenced as the first-listed or principal diagnosis.




When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
